---
title: 'FIFA 19 Player Data Analysis'
author: "STAT 420, Summer 2020, Shuo Yan, Chutong Xiao, Connor Ng"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
library(knitr)
opts_chunk$set(cache = TRUE, autodep = TRUE)
```

```{r}
birthday = 19980327
set.seed(birthday)
```

## Introduction

Everyone of us can list our favorite football players; however, we as amateur audiences seldom know why some football players are so famous and their transfer values are so high. Hopefully, through this study we will be able to understand the standard of player evaluation. Are certain traits in football players more beneficial than others? Is pure physical ability enough to be thought to be as good as Lionel Messi or Cristiano Ronaldo or is there something more?

In this study, we will use the `FIFA_19.csv` dataset to explore the relationship between the overall ratings and other attributes of the players. We aim the find the best model that help determine the rating of a palyer based on the existing data.

The variables in the dataset are:

- `ID` 
- `Name` - Fullname of the football player
- `Age` - in years
- `Photo` - Link to the profile photo 
- `Nationality` 
- `Flag` - Link to the national flag photo 
- `Overall` - The overall rating, integer
- `Potential` - The potential rating, integer
- `Club`
- `Club.logo` - Link to the Logo image

- `Value` - in M EUR
- `Wage` - in K EUR
- `Release.Cause` - M EUR

- `Special` - 
- `Preferred.Foot` - right or left
- `International.Reputation` - 
- `Weak.Foot` - 
- `Skill.Moves` -
- `Work.Rate` - 
- `Body.Type` - Lean, normal or stocky
- `Real.Face` - Whether the player has a real face (Yes or No)
- `Position` 
- `Jersey.Number` - Integer that identify the position
- `Joined` - The year when the player joined FIFA
- `Contract.Valid.Until` - Contract end-dates in years
- `Height` - in feet
- `Weight` - in lbs
- `LS` - Left Striker
- `ST` - Striker
- `RS` - Right Striker
- `LW` - Left Wing
- `RW` - Right Wing
- `LF` - Left Forward
- `CF` - Centre Forward
- `RF` - Right Forward
- `LAM` - Left Attacking Midfielder
- `CAM` - Central Attacking Midfielder
- `RAM` - Right Attacking Midfielder
- `LM` - Left Midfielder
- `RM` - Right Midfielder
- `LDM` - Left Defensive Midfielder
- `CDM` - Central Defensive Midfielder
- `RDM` - Right Defensive Midfielder
- `LB` - Left Back (##to delete## notice: we do not have right back(RB) in the dataset)
- `LCM` - Left Centre Midfielder
- `CM` - Central Midfielder
- `RCM` - Right Centre Midfielder
- `LWB` - Left Wing Back
- `RWB` - Right Wing Back

- Attacking: `Crossing`, `Finishing`, `HeadingAccuracy`, `ShortPassing`, `Volleys`

- Skill: `Dribbling`, `Curve`, `FKAccuracy`, `LongPassing`, `BallControl`

- Movement: `Acceleration`, `SprintSpeed`, `Agility`, `Reactions`, `Balance`

- Power: `ShotPower`, `Jumping`, `Stamina`, `Strength`, `LongShots`

- Mentality: `Aggression`, `Interceptions`, `Positioning`, `Vision`, `Penalties`, `Composure`                 
- Defending: `Marking`, `StandingTackle`, `SlidingTackle`

- GoalKeeping: `GKDiving`, `GKHandling`, `GKKicking`, `GKPositioning`,`GKReflexes`






## Methods

### Data Preprocessing

```{r evaluate = FALSE, warning = FALSE}
raw_data = read.csv("/Users/shuoyan/Documents/stat420/finalProject/fifa18-all-player-statistics-master/2019/data.csv")

library(dplyr)
#remove obviously unuseful variables
file = select(raw_data,-c(X,ID, Photo, Flag, Club.Logo, Loaned.From))

#convert columns LS to RB to numeric values
# take the values with plus sign and without plus sign
position_raw = file[, 23:48]
original = matrix(0, nrow(position_raw), ncol(position_raw))
after = matrix(0, nrow(position_raw), ncol(position_raw)) #######THIS IS LS TO RB WITH PLUS SIGN!!!
for(i in 1:nrow(position_raw)){
  for(j in 1:ncol(position_raw)){
    str = as.character(position_raw[i, j])
    if(str == ""){
      original[i, j] = NA
      after[i, j] = NA
    } else{
      str = strsplit(str, split = "+")[[1]]
      original[i,j] = as.numeric(paste(str[1], str[2], sep = ""))
      after[i,j] = original[i] + as.numeric(str[4])
    }
  }
}

file[, 23:48] = original
file = na.omit(file)

#convert Value into numeric
file$Value = as.character(file$Value)
#remove euro sign and million
temp = substring(strsplit(file$Value, split = "M"), 2) ##########NOTE: this value is in Million
file$Value = as.numeric(temp)
#remove NA
anyNA(file$Value)
if(anyNA(file$Value)){
  file = na.omit(file)
}

#convert wage into numeric
file$Wage = as.character(file$Wage)
#remove euto sign and K 
temp = substring(strsplit(file$Wage, split = "K"), 2) ###########NOTE: this value is in K
file$Wage = as.numeric(temp)
#remove NA
anyNA(file$Wage)


#convert release clause into numeric
file$Release.Clause = as.character(file$Release.Clause)
#remove euro sign and M
temp = substring(strsplit(file$Release.Clause, split = "M"), 2) #########NOTE: this value is in M
file$Release.Clause = as.numeric(temp)
#remove NA
if(anyNA(file$Release.Clause)){
  file = na.omit(file)
}

#convert weight from factor to numeric
#blank is missing data
file = subset(file, class(file$Weight) != as.factor(""))
file$Weight = substr(file$Weight, 1, 3)

#convert height to numeric
#blank is missing data
file = subset(file, class(file$Height) != as.factor(""))
file$Height = as.numeric(gsub("\'", ".", file$Height)) * 0.3048 ########Note: in meters

#convert weight to numeric
file$Weight = as.numeric(file$Weight)

#check if columns are their expected types and remove NA

data.frame(colnames(file))
numeric_index = c(4, 2, 5, 7, 8, 9, 11, 12, 13, 21:83)
factor_index = c(1, 3, 6, 10, 14:20)
for(i in numeric_index){
  #print(class(file[, i]))
  file[,i] = na.omit(file[,i])
}

for(i in factor_index){
  #print(class(file[,i]))
  file = subset(file, class(file[, i])!= as.factor(""))
}

#change jersey number to categorical
file$Jersey.Number = as.factor(file$Jersey.Number)

file = na.omit(file)
```


### Data Analysis
#### Check for Collinearity of Numeric parameters

- collinearity of physical scorting and Overall Score

Because a full model crashes rstudio, we first analyze some very important parameters - the scoring of player's physical attributes.

```{r}
selection_model = lm(Overall ~ Age + Height + Weight + Weak.Foot + Skill.Moves + Crossing + Finishing + HeadingAccuracy + ShortPassing + Volleys + Dribbling + Curve + FKAccuracy + LongPassing + BallControl + Acceleration + SprintSpeed + Agility + Reactions + Balance + ShotPower + Jumping + Stamina + Strength + LongShots + Aggression + Interceptions + Positioning + Vision + Penalties + Composure +  Marking + StandingTackle + SlidingTackle + GKDiving + GKHandling + GKKicking + GKPositioning + GKReflexes, data = file)

library(faraway)

vif(selection_model)
```
To ensure that the final model doesn't have collinearity issues, we would select parameters with vif less than 5. Therefore we select parameters Age, Height, Weight, Weak.Foot, Skill.Moves, Crossing, HeadingAccuracy, ShortPassing, Volleys, Curve, FKAccuracy, LongPassing, Agility, Reactions, Balance, ShotPower, Jumping, Stamina, Strength, Aggression, Vision, Penalties, Composure, GKDiving, GKHandling, GKKicking, GKPositioning, GKReflexes.

- correlation between Overall score and position scores

```{r}
selection_model = lm(Overall ~ ., data = file[, c(4, 23 : 48)])
vif(selection_model)
```

We see that all of the parameters have large vifs. To avoid collinearity issues, we would use none of them.

Therefore, we would use the above parameters and select remaining categorical parameters to fit models.

#### Check for Influential Points

```{r}
check_influential_model = lm(Overall ~ Age + Height + Weight + Weak.Foot + Skill.Moves + Crossing + HeadingAccuracy + ShortPassing + Volleys + Curve + FKAccuracy + LongPassing + Agility + Reactions + Balance + ShotPower + Jumping + Stamina + Strength  + Aggression + Vision + Penalties + Composure + GKDiving + GKHandling + GKKicking + GKPositioning + GKReflexes, data = file)

which(cooks.distance(check_influential_model) > 4 / length(cooks.distance(check_influential_model)))
```

We wouldn't consider removing any influential points because there are many of them. Also consider that talents vary, it's unreasonable to remove them.

We have finished analyzing data and variables. Now we can fit models!

### Model Fitting
